Modelling BCG vaccination in the UK

### 2019-07-11 rmds/02-background.Rmd
updated background based on sup comments

Tuberculosis (TB) is thought to infect over 1.7 billion people globally, of which 5-15% will develop active TB in their lifetime.[@TheWorldHealthOrganization:2018va] Of this number around 10% are likely to die from TB or TB related causes. TB is preventable and curable, but the majority of cases occur in less economically developed countries and are never diagnosed.[@TheWorldHealthOrganization:2018va] In the following section, the natural history of TB, the risk factors, the treatment, the global impact, and the impact in England and Wales are explored.

@@

TB is primarily a respiratory disease (pulmonary TB) caused by the bacterium Mycobacterium TB, although it can also affect other parts of the body (extra-pulmonary TB). TB spreads via airborne droplets that are expelled when individuals with active pulmonary TB cough. After an infection with TB, 5-10% of individuals develop primary disease within 1-2 years of exposure. Children are more likely to develop active disease and to develop it more quickly than adults.[@TheWorldHealthOrganization:2018va] The majority of individuals then enter a latent stage in which they passively carry TB mycobacterium. Reactivation of bacilli can then occur many years later due to a loss of immune control.[@Gideon2011a]

Both active and latent TB cases represent a range of diverse individual states. Pulmonary cases are typically responsible for the vast majority of transmission.[@Sepkowitz1996] Latent cases may have completely cleared the bacterium or be asymptotically carrying reproducing active TB bacterium.[@Gideon2011a] Adolescents have the highest risk of developing active TB, usually in the form of pulmonary TB.[@TheWorldHealthOrganization:2018va] The risk of developing pulmonary TB, versus extra-pulmonary TB, varies with age. For instance, younger children are more likely to develop pulmonary TB.[@TheWorldHealthOrganization:2018va]

@@

TB has been associated with several risk factors, the most common of which is HIV. HIV increases the rate of activation by 20-fold and TB is the most common cause of AIDS-related death.[@Rottenberg2012] Increased risk of TB can also be the result of other medical conditions, such as diabetes, or lifestyle and environmental factors. These include smoking, low socioeconomic status, high density living, homelessness, incarceration, and drug use.[@Bhatti1995; @Narasimhan2013; @Story2007]

@@

Treatment for TB consists of a six month course of multiple antibiotics (see Table \@ref(tab:tab-interventions)). These usually consist of isoniazid, rifampicin,  pyrazinamide and ethambutol (known as first line drugs). If the disease is resistant to treatment with the first line drugs then second line drugs such as aminoglycosides, fluoroquinolones, and cycloserine are employed. The side effects for these drugs are generally far more severe and the treatment regime is longer, typically 12-24 months. The World Health Organisation now recommends the use of the Directly Observed Treatment short-course (DOTS), which focuses on 5 action points.[@WHOTB2016] These are:

@@

```{r tab-interventions, echo = FALSE, message = FALSE, dpi = 500}
tb_interventions_timeline %>% 
  set_names(c("Year",
              "Intervention",
              "Type",
              "Line",
              "Detail")) %>% 
  knitr::kable(
    caption.short = "A timeline of interventions against TB",
    caption = "A timeline of interventions against TB. Antibiotics used to treat TB are commonly given together, with those with the fewest side effects given first. Second line antibiotics are then used if the initial treatment fails or tests show the strain is multiply drug resistant. BCG - Bacillus Calmette–Guérin; TB – Tuberculosis; MRSA - Methicillin-resistant Staphylococcus aureus; DOTS - Directly Observed Treatment Short-course", booktabs = TRUE, longtable = TRUE, escape = FALSE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(5, width = "10cm") %>% 
  kable_styling(latex_options = "hold_position") %>% 
  landscape()
```

@@

TB is a global disease with an estimated 10.4 million new cases in 2016, of which 4.3 million were missed by health systems.[@TheWorldHealthOrganization:2018va]  Global incidence rates have decreased year-on-year since the early 2000's, with an average year on year decrease of 2.9%. However, global TB incidence remains above 134 per 100,000 population (Figure \@ref(fig:global-regional-tb-summary)). On a regional level, incidence rates vary with Africa and South-East Asia having a greater concentration of cases. In the Eastern Mediterranean incidence rates have remained relatively stable over the last 10 years.

```{r global-regional-tb-summary, fig.cap = "TB incidence rates (per 100,000) by region and globally from 2000 until 2017. Globally incidence rates have been declining since the early 2000's but this decline varies with region. Plot produced using getTBinR", fig.scap = "TB incidence rates (per 100,000) by region and globally from 2000 until 2017.", out.extra = ""}
plot_tb_burden_summary(conf = NULL, verbose = FALSE)
```

Regional incidence rates only tell part of the story as TB incidence rates vary significantly within regions. Six countries: India, Indonesia, China, Nigeria, Pakistan, and South Africa account for 60% of new cases. India, Indonesia, and Nigeria comprise nearly half of the gap between incident and notified cases.[@WHOTB2016] Figure \@ref(fig:map-global-tb) shows both regional similarities and countries, like Mongolia, that stand out as having higher TB incidence rates than surrounding countries.

```{r map-global-tb, fig.cap = "Global map of country level TB incidence rates (per 100,000 population) in 2017. Note the clustering of countries with high incidence rates in southern and central Africa and southern Asia. Map produced using getTBinR", fig.scap = "Global map of country level TB incidence rates (per 100,000 population) in 201", fig.height = 5}
map_tb_burden(verbose = FALSE)
```

TB remains one of the top 10 causes death worldwide, leading to 1.7 million deaths in 2016 alone.[@TheWorldHealthOrganization:2018va]  Whilst the absolute number of deaths due to TB has fallen since 2000, the average global rate of decline in TB mortality rates was only 2.9% between 2000-2016. However, unlike the trend observed for incidence rates, the year-on-year decline of TB mortality rates has remained consistent in all regions (Figure \@ref(fig:global-regional-mortality-summary)). Several regions, including Africa and Europe, have seen TB mortality rates fall to below 50% of those in 2000.

```{r global-regional-mortality-summary, fig.scap =  "TB mortality rates (per 100,000) by region and globally from 2000 until 2017.", fig.cap = "TB mortality rates (per 100,000) by region and globally from 2000 until 2017. Mortality rates from TB have been falling in all regions since 2000. Plot produced using getTBinR", out.extra = ""}
plot_tb_burden_summary(conf = NULL, verbose = FALSE, metric = "e_mort_100k", 
                       denom = "e_inc_num", facet = "Area", legend = "none", 
                       scales = "free_y") 
```

@@

```{r map-global-tb-with-hiv, fig.scap = "Global map of estimated HIV in incidence TB (percent) in 2017.", fig.cap = "Global map of estimated HIV in incidence TB (percent) in 2017. Note that high percentage of TB cases with HIV in sub-saharan Africa. Map produced using getTBinR", fig.height = 5}
map_tb_burden(metric = "e_tbhiv_prct", verbose = FALSE)
```

Multi-drug-resistant TB (MDR-TB), which is defined as being resistant to at least isoniazid and rifampin, made up 4.6% of all new TB cases in 2015 (480,000). It can be acquired both through treatment failure and through transmission. Treatment requires the use of second line antibiotics, which often have more severe side effects and are more likely to fail, with only 52% successfully treated globally compared to 83% for drug susceptible TB.[@WHOTB2016] As for HIV co-infection, drug resistance is globally heterogeneous with some regions, like countries in the former USSR, having a much higher proportion of drug resistant cases. Figure \@ref(fig:map-global-tb-with-rr) shows the country level proportion of cases with at least rifampicin resistance and highlights the higher level of rifampicin resistance in countries formerly in the USSR. 88% of rifampicin cases in Russia in 2017 also had MDR-TB, which is comparable to the global median of 83% (Table \@ref(tab:prop-rr-with-mdr-tab)). Across all regions Europe had the highest median percentage of rifampicin cases with MDR-TB (100%), with Africa having the lowest (67%). This variation may indicate areas in which drug resistance is developing (regions with low proportions of MDR to rifampicin resistance), from regions in which MDR-TB results from transmission or importation (regions with high proportions of MDR to rifampicin resistance).

```{r map-global-tb-with-rr, fig.scap = "Global map of the estimated percentage of new TB cases with rifampicin resistance (percent) in 2017.", fig.cap = "Global map of the estimated percentage of new TB cases with rifampicin resistance (percent) in 2017. Not that a far higher percentage of TB cases have rifampicin resistance in the former Soviet Union that in the rest of the world. Map produced using getTBinR (see \\@ref(getTBinR) for details)", fig.height = 5}
map_tb_burden(metric = "e_rr_pct_new", verbose = FALSE)
```

```{r prop-rr-with-mdr-tab, warning = FALSE}
summarise_tb_burden(countries = "Russia", 
                    metric = "e_mdr_pct_rr_new", verbose = FALSE,
                    year = 2017, stat = "median") %>% 
  mutate_at(.vars = c("e_mdr_pct_rr_new", "e_mdr_pct_rr_new_lo", "e_mdr_pct_rr_new_hi"),
            funs(ifelse(. > 100, 100, .))) %>% 
  mutate(`Median percentage of RR* cases with MDR** (95% IQR***)` = 
           pretty_ci(e_mdr_pct_rr_new,
                     e_mdr_pct_rr_new_lo, 
                     e_mdr_pct_rr_new_hi)) %>% 
  mutate(`Median Proportion of RR* cases with MDR** (95% IQR***)` = 
           ifelse(area == "Russian Federation", paste(pretty_round(e_mdr_pct_rr_new), "***"),
                  as.character(`Median percentage of RR* cases with MDR** (95% IQR***)`)
                  )) %>% 
  select(Area = area, `Median percentage of RR* cases with MDR** (95% IQR***)`) %>% 
  knitr::kable(
               caption = "Percentage (\\%) of rifampicin resistant TB cases that have multi-drug resistant TB in Russia and regional medians, with interquartile ranges.",
        booktabs = TRUE) %>% 
  add_footnote(label = c("* Rifampicin resistance", "** Multi-Drug Resistant TB", "*** Interquartile Range"), notation = "none") %>% 
  kable_styling(latex_options = "hold_position") %>% 
  {.}
```

@@

TB incidence in England and Wales has decreased dramatically from a century ago (Figure \@ref(fig:plot-incidence), or see [http://www.seabbs.co.uk/shiny/TB_England_Wales](http://www.seabbs.co.uk/shiny/TB_England_Wales) for an interactive dashboard). However, in the past several decades, incidence rates first stabilised and have since increased since their lowest point in the 1990's. In 2000 there were 6044 notified TB cases in England, increasing to a maximum of 8280 notified TB cases in 2011. Since then, notifications have declined year on year.[@PHE2016a] Figure \@ref(fig:plot-incidence) includes the interventions discussed above (Table \@ref(tab:tab-interventions)) and indicates that the introduction of several antibiotics and BCG vaccination in the 1950s may have led to an extended decrease in incidence.

```{r plot-incidence, echo = FALSE, warning = FALSE, message = FALSE, dpi = 300, fig.height = 10, fig.width = 10, fig.scap = "TB notifications in England and Wales from 1913 to 2017, stratified initially by respiratory/non-respiratory status and from 1982 by pulmonary/non-pulmonary TB.", fig.cap = "TB notifications in England and Wales from 1913 to 2017, stratified initially by respiratory/non-respiratory status and from 1982 by pulmonary/non-pulmonary TB. Interventions are highlighted with vertical lines, with linetype denoting the type of intervention, more information on each intervention is available in the corresponding table. Figure produced using tbinenglanddataclean (https://www.samabbott.co.uk/tbinenglanddataclean/)", out.extra = ""}
plot_historic_tb_ew() + theme(legend.position = "top")
```

@@

TB incidence in England and Wales is highly heterogeneous with over 70% of cases occurring in the non-UK born population. Incidence rates in the non-UK born (49.4 per 100,000, in 2016) are 15 times higher than in the UK born population (3.2 (95% CI 3.0-3.3) per
100,000, in 2016).[@PHE2017] The age distribution of cases in the UK born and non-UK born populations differ, with the UK born population having a relatively uniform distribution. Meanwhile, the non-UK born have higher incidence rates in those aged 80 years and older (69.3 per 100,000 in 2016), those aged 75 to 79 years (62.9 per 100,000 in 2016) and those aged 25-29 years old (61.6 per 100,000 in 2016) [@PHE2017]. In the non-UK born, the majority of cases occur amongst those who have lived in the UK for at least 6 years (63%) - this has increased year on year since 2010 (49%).[@PHE2017] However, in 2016, 23.3% (420/1,800) of non-UK born cases had traveled outside the UK, with the majority returning to their country of origin. Incidence rates in the UK born are between 3 and 14 times higher in non-White ethnic groups compared to the White ethnic groups.[@PHE2017]

The majority of cases occur in urban areas. London alone accounts for 39% of cases, with an incidence rate of 25.1 (per 100,000; 95% CI 24.1-26.2; in 2016).[@PHE2017] England has few cases of MDR-TB cases, with only 68 cases recorded in 2016. Similarly the number of co-infections with HIV is low with only 3.8% of cases in 2015 having HIV - the majority of these cases were born in countries with high HIV prevalence. In 2016, 11.1% of TB cases in 2016 had at least one social risk factor, compared with 11.7% in 2015.[@PHE2017] In general cases with social risk factors are more likely to have drug resistant TB, worse TB outcomes, and to be lost to follow up. [@PHE2017] Amongst cases who were of working age in 2016, with a known occupation; 35.2% (1,491/4,240) were not in education or employment, 10.2% (432) were either studying or working in education; and 7.1% (304) were healthcare workers.[@PHE2017]

@@

Figure \@ref(fig:plot-prop-pul) shows that since the 1980s the proportion of extra-pulmonary TB compared to pulmonary TB has increased from `r prop_extra_pil_1982` in 1982 to `r prop_extra_pil_2016` in 2016. This may be attributed to the age distribution of TB cases changing as different age groups are more likely to progress to pulmonary vs extra-pulmonary TB. It may also be related to the increase of non-UK born cases as a higher proportion of non-UK born cases have extra-pulmonary disease only (51.4%, 2,103/4,089, in 2016), compared to UK born cases (31.9%, 467/1,465, in 2016).[@PHE2017] For more details on TB in England, see Chapter \@ref(data) and the Public Health England 2017 and 2018 TB reports from which the summary data discussed above was extracted.[@PHE2017; @PHE2018]

```{r plot-prop-pul, echo = FALSE, message = FALSE, warning = FALSE, dpi = 300, fig.height = 10, fig.width = 10, fig.scap = "From 1913 until 1981 the figure shows the proportion respiratory vs. non-respiratory cases and from 1982 it shows the proportion of pulmonary vs. non-pulmonary TB.", fig.cap = "From 1913 until 1981 the figure shows the proportion respiratory vs. non-respiratory cases and from 1982 it shows the proportion of pulmonary vs. non-pulmonary TB. Figure produced using tbinenglanddataclean (https://www.samabbott.co.uk/tbinenglanddataclean/)"}
plot_historic_prop_pul_tb() + theme(legend.position = "top")
```

@@

Until recently little attention has been given to any additional effects of BCG vaccination.[@Fine2005a; @Teo2006] However, there is now some evidence that BCG vaccination induces innate immune responses that may provide non-specific protection[@Kleinnijenhuis2012] and reduce all-cause neonatal mortality.[@Garly2003; @Higgins] There is some evidence that this reduction in all-cause mortality extends later in life.[@Rieckmann2016] Additionally, BCG vaccination may improve outcomes for individuals with active TB disease. TB patients with BCG scars have been found to respond better to treatment with earlier sputum smear conversion.[@Jeremiah2010] There is also evidence to support an association between BCG vaccination and reduced TB[@Abubakar2013] mortality. The evidence for additional effects of BCG vaccination on outcomes in individuals with notified TB, in England, is explored further in Chapter \@ref(beneficial-bcg-out).

@@

The BCG vaccine is one of the mostly widely-used vaccines worldwide, with approximately 100m doses given annually.[@The2004] However, due to the variable estimates of BCG efficacy, vaccination has been controversial since its development. The World Health Organisation (WHO) recommends vaccination for all neonates as early as possible after birth in high burden settings. Vaccination in low burden settings is dependent on the country specific epidemiology of TB.[@WHO2017; @TheWorldHealthOrganization:2018va] This recommendation is based on the strong evidence that the BCG is highly protective in children,[@Rodrigues1993; @Colditz1994] whilst it's effectiveness has been shown to vary with latitude when given later in life.[@Mangtani2014] Historically, different strategies have been utilized worldwide. This includes universal vaccination of those at most risk of on-wards transmission and high-risk group vaccination targeting either neonates or children.[@Abubakar2013]

@@

Multiple replacement vaccines are currently in clinical trials.[@Zwerling2011a; @Schrager:2018ip] Vaccine candidates include both live and sub-unit vaccines. Many of these candidate vaccines serve as a boost to the BCG vaccine, with the BCG vaccine being administered prior to the candidate vaccine.[@Zwerling2011a] Several BCG replacements are also being trialed, both based on alternative methods of attenuating TB mycobacteria and using other approaches.[@Schrager:2018ip] However, in the short-to-midterm it is unlikely that a new vaccine will replace the BCG vaccine. This means that it's optimal usage is as important as ever.

>>>>

Tuberculosis (TB) is thought to infect over 1.7 billion people globally, of which 5-15% will develop active TB in their lifetime.[@Houben:2016jp] Of this number around 10% are likely to die from TB or TB related causes.[@TheWorldHealthOrganization:2018va] TB is preventable and curable, but the majority of cases occur in less economically developed countries and are never diagnosed.[@TheWorldHealthOrganization:2018va] In the following section, the natural history of TB, the risk factors, the treatment, the global impact, and the impact in England and Wales are explored.

@@

TB is primarily a respiratory disease (pulmonary TB) caused by the bacterium Mycobacterium TB, although it can also affect other parts of the body (extra-pulmonary TB). TB spreads via airborne droplets that are expelled when individuals with active pulmonary TB cough. After infection with TB, 5-10% of individuals develop primary disease within 1-2 years of exposure. Children are more likely to develop active disease and to develop it more quickly than adults.[@TheWorldHealthOrganization:2018va] The remaining 90-95% of individuals then enter a latent stage in which they passively carry TB mycobacterium. Reactivation of bacilli can then occur many years later due to a loss of immune control.[@Gideon2011a]

Both active and latent TB represent a range of diverse individual states. Pulmonary cases are typically responsible for the vast majority of transmission.[@Sepkowitz1996] Latent cases may completely clear the bacterium or be asymptotically carrying reproducing active TB bacterium.[@Gideon2011a] Adolescents have the highest risk of developing active TB, usually in the form of pulmonary TB.[@TheWorldHealthOrganization:2018va] The risk of developing pulmonary TB, versus extra-pulmonary TB, varies with age. For instance, younger children are more likely to develop pulmonary TB.[@TheWorldHealthOrganization:2018va]

@@

TB has been associated with several risk factors, the most common of which is human immunodeficiency viruse (HIV). HIV increases the rate of activation by 20-fold and TB is the most common cause of AIDS-related death.[@Rottenberg2012] Increased risk of TB can also be the result of other medical conditions, such as diabetes, or lifestyle and environmental factors. These include smoking, low socioeconomic status, high density living, homelessness, incarceration, and drug use.[@Bhatti1995; @Narasimhan2013; @Story2007]

@@

Treatment for TB consists of a six month course of multiple antibiotics (see Table \@ref(tab:tab-interventions)). These usually consist of isoniazid, rifampicin,  pyrazinamide and ethambutol (known as first line drugs). If the disease is resistant to treatment with the first line drugs then second line drugs such as aminoglycosides, fluoroquinolones, and cycloserine are employed. The side effects for these drugs are generally far more severe and the treatment regime is longer, typically 12-24 months. The World Health Organization (WHO) now recommends the use of the Directly Observed Treatment short-course (DOTS), which focuses on 5 action points.[@WHOTB2016] These are:

@@

```{r tab-interventions, echo = FALSE, message = FALSE, dpi = 500}
tb_interventions_timeline %>% 
  set_names(c("Year",
              "Intervention",
              "Type",
              "Line",
              "Detail")) %>% 
  knitr::kable(
    caption.short = "A timeline of interventions against TB",
    caption = "A timeline of interventions against TB. Antibiotics used to treat TB are commonly given together, with those with the fewest side effects given first. Second line antibiotics are then used if the initial treatment fails or tests show the strain is multiply drug resistant. BCG - Bacillus Calmette–Guérin; TB – Tuberculosis; MRSA - Methicillin-resistant *Staphylococcus aureus*; DOTS - Directly Observed Treatment Short-course", booktabs = TRUE, longtable = TRUE, escape = FALSE) %>% 
  kable_styling(font_size = 8, latex_options = c("repeat_header")) %>% 
  column_spec(5, width = "10cm") %>% 
  kable_styling(latex_options = "hold_position") %>% 
  landscape()
```

@@

TB is a global disease with an estimated 10.4 million new cases in 2016,[@TheWorldHealthOrganization:2018va] of which 4.3 million were missed by health systems.[@WHO:2018]  Global incidence rates have decreased year-on-year since the early 2000s, with an average year on year decrease of 2.9%. However, global TB incidence remains above 134 per 100,000 population (Figure \@ref(fig:global-regional-tb-summary)). On a regional level, incidence rates vary with Africa and South-East Asia having a greater concentration of cases. In the Eastern Mediterranean incidence rates have remained relatively stable over the last 10 years.

```{r global-regional-tb-summary, fig.cap = "TB incidence rates (per 100,000) by region and globally from 2000 until 2017. Globally incidence rates have been declining since the early 2000s but this decline varies with region. Plot produced using getTBinR", fig.scap = "TB incidence rates (per 100,000) by region and globally from 2000 until 2017.", out.extra = ""}
plot_tb_burden_summary(conf = NULL, verbose = FALSE, viridis_end = 1)
```

Regional incidence rates only tell part of the story as TB incidence rates vary significantly within regions. Six countries: India, Indonesia, China, Nigeria, Pakistan, and South Africa account for 60% of new cases. India, Indonesia, and Nigeria are thought to be responsible for nearly half of all undiagnosed cases.[@WHOTB2016] Figure \@ref(fig:map-global-tb) shows both regional similarities and countries, like Mongolia, that stand out as having higher TB incidence rates than surrounding countries.

```{r map-global-tb, fig.cap = "Global map of country level TB incidence rates (per 100,000 population) in 2017. Note the clustering of countries with high incidence rates in southern and central Africa and southern Asia. Incidence rates in the legend refer to the lower bound for each colour. Map produced using getTBinR", fig.scap = "Global map of country level TB incidence rates (per 100,000 population) in 201", fig.height = 6}
map_tb_burden(verbose = FALSE)
```

TB remains one of the top 10 causes death worldwide, leading to 1.7 million deaths in 2016 alone.[@TheWorldHealthOrganization:2018va]  Whilst the absolute number of deaths due to TB has fallen since 2000, the average annual global rate of decline in TB mortality rates was only 2.9% between 2000-2016. However, unlike the trend observed for incidence rates, the year-on-year decline of TB mortality rates has remained consistent in all regions (Figure \@ref(fig:global-regional-mortality-summary)). Several regions, including Africa and Europe, have seen TB mortality rates fall to below 50% of those in 2000.

```{r global-regional-mortality-summary, fig.scap =  "TB mortality rates (per 100,000 cases) by region and globally from 2000 until 2017.", fig.cap = "TB mortality rates (per 100,000 cases) by region and globally from 2000 until 2017. Mortality rates from TB have been falling in all regions since 2000. Plot produced using getTBinR", out.extra = ""}
plot_tb_burden_summary(conf = NULL, verbose = FALSE, metric = "e_mort_100k", 
                       denom = "e_inc_num", facet = "Area", legend = "none", 
                       scales = "free_y") 
```

@@

```{r map-global-tb-with-hiv, fig.scap = "Global map of estimated HIV in incidence TB (percent) in 2017.", fig.cap = "Global map of estimated HIV in incidence TB (percent) in 2017. Note the high percentage of TB cases with HIV in sub-saharan Africa. The percentages of HIV in incident TB in the legend refer to the lower bound for each colour. Map produced using getTBinR", fig.height = 6}
map_tb_burden(metric = "e_tbhiv_prct", verbose = FALSE)
```

Multi-drug-resistant TB (MDR-TB), which is defined as being resistant to at least isoniazid and rifampin, made up 4.6% of all new TB cases globally in 2015 (480,000). It can be acquired both through treatment failure and through transmission. Treatment requires the use of second line antibiotics, which often have more severe side effects and are more likely to fail, with only 52% successfully treated globally compared to 83% for drug susceptible TB.[@WHOTB2016] As for HIV co-infection, drug resistance is globally heterogeneous with some regions, like countries in the former USSR, having a much higher proportion of drug resistant cases. Figure \@ref(fig:map-global-tb-with-rr) shows the country level proportion of cases with at least rifampicin resistance and highlights the higher level of rifampicin resistance in countries formerly in the USSR. 88% of rifampicin cases in Russia in 2017 also had MDR-TB, which is comparable to the global median of 83% (Table \@ref(tab:prop-rr-with-mdr-tab)). Across all regions Europe had the highest median percentage of rifampicin cases with MDR-TB (100%), with Africa having the lowest (67%). This variation may indicate areas in which drug resistance is developing (regions with low proportions of MDR to rifampicin resistance), from regions in which MDR-TB results from transmission or importation (regions with high proportions of MDR to rifampicin resistance).

```{r map-global-tb-with-rr, fig.scap = "Global map of the estimated percentage of new TB cases with rifampicin resistance (percent) in 2017.", fig.cap = "Global map of the estimated percentage of new TB cases with rifampicin resistance (percent) in 2017. Note that a far higher percentage of TB cases have rifampicin resistance in the former Soviet Union that in the rest of the world. The percentages of rifampicin resistances in incident TB in the legend refer to the lower bound for each colour. Map produced using getTBinR (see \\@ref(getTBinR) for details)", fig.height = 6}
map_tb_burden(metric = "e_rr_pct_new", verbose = FALSE)
```

```{r prop-rr-with-mdr-tab, warning = FALSE}
summarise_tb_burden(countries = "Russia", 
                    metric = "e_mdr_pct_rr_new", verbose = FALSE,
                    year = 2017, stat = "median") %>% 
  mutate_at(.vars = c("e_mdr_pct_rr_new", "e_mdr_pct_rr_new_lo", "e_mdr_pct_rr_new_hi"),
            funs(ifelse(. > 100, 100, .))) %>% 
  arrange(desc(area)) %>% 
  mutate(`Median percentage of RR* cases with MDR** (95% IQR***)` = 
           pretty_ci(e_mdr_pct_rr_new,
                     e_mdr_pct_rr_new_lo, 
                     e_mdr_pct_rr_new_hi, digits = 1)) %>% 
  mutate(`Median Proportion of RR* cases with MDR** (95% IQR***)` = 
           ifelse(area == "Russian Federation", paste(pretty_round(e_mdr_pct_rr_new), "***"),
                  as.character(`Median percentage of RR* cases with MDR** (95% IQR***)`)
                  )) %>% 
  select(Area = area, `Median percentage of RR* cases with MDR** (95% IQR***)`) %>% 
  knitr::kable(
               caption = "Percentage (\\%) of rifampicin resistant TB cases that have multi-drug resistant TB in Russia and regional medians, with interquartile ranges.",
        booktabs = TRUE) %>% 
  add_footnote(label = c("* Rifampicin resistance", "** Multi-Drug Resistant TB", "*** Interquartile Range"), notation = "none") %>% 
  kable_styling(latex_options = "hold_position") %>% 
  {.}
```

@@

TB incidence in England and Wales has decreased dramatically from a century ago (Figure \@ref(fig:plot-incidence)^[or see [http://www.seabbs.co.uk/shiny/TB_England_Wales](http://www.seabbs.co.uk/shiny/TB_England_Wales]) for an interactive dashboard). However, in the past several decades, incidence rates first stabilised and have since increased since their lowest point in the 1990s. In 2000, there were 6044 notified TB cases in England, increasing to a maximum of 8280 notified TB cases in 2011. Since then, notifications have declined year on year.[@PHE2016a] Figure \@ref(fig:plot-incidence) includes the interventions discussed above (Table \@ref(tab:tab-interventions)) and indicates that the introduction of several antibiotics and BCG vaccination in the 1950s may have led to an extended decrease in incidence.

```{r plot-incidence, echo = FALSE, warning = FALSE, message = FALSE, dpi = 300, fig.height = 12, fig.width = 12, fig.scap = "TB notifications in England and Wales from 1913 to 2017, stratified initially by respiratory/non-respiratory status and from 1982 by pulmonary/non-pulmonary TB.", fig.cap = "TB notifications in England and Wales from 1913 to 2017, stratified initially by respiratory/non-respiratory status and from 1982 by pulmonary/non-pulmonary TB. Interventions are highlighted with vertical lines, with linetype denoting the type of intervention, more information on each intervention is available in the corresponding table. Figure produced using `tbinenglanddataclean`", out.extra = ""}
plot_historic_tb_ew() + theme(legend.position = "top")
```

@@

TB incidence in England and Wales is highly heterogeneous with over 70% of cases occurring in the non-UK born population. Incidence rates in the non-UK born (49.4 per 100,000, in 2016) are 15 times higher than in the UK born population (3.2 per
100,000, in 2016).[@PHE2017] The age distribution of cases in the UK born and non-UK born populations differ, with the UK born population having a relatively uniform distribution. Meanwhile, the non-UK born have higher incidence rates in those aged 80 years and older (69.3 per 100,000 in 2016), those aged 75 to 79 years (62.9 per 100,000 in 2016) and those aged 25-29 years old (61.6 per 100,000 in 2016) [@PHE2017]. In the non-UK born, the majority of cases occur amongst those who have lived in the UK for at least 6 years (63%) - this has increased year on year since 2010 (when it was 49%).[@PHE2017] However, in 2016, 23.3% (420/1,800) of non-UK born cases had traveled outside the UK, with the majority returning to their country of origin. Incidence rates in the UK born are between 3 and 14 times higher in non-White ethnic groups compared to the White ethnic groups.[@PHE2017]

The majority of cases occur in urban areas. London alone accounts for 39% of cases, with an incidence rate of 25.1 (per 100,000, in 2016).[@PHE2017] England has few cases of MDR-TB cases, with only 68 cases recorded in 2016. Similarly the number of co-infections with HIV is low with only 3.8% of cases in 2015 having HIV - the majority of these cases were born in countries with high HIV prevalence. In 2016, 11.1% of TB cases in 2016 had at least one social risk factor, compared with 11.7% in 2015.[@PHE2017] In general cases with social risk factors are more likely to have drug resistant TB, worse TB outcomes, and to be lost to follow up. [@PHE2017] Amongst cases who were of working age in 2016, with a known occupation; 35.2% (1,491/4,240) were not in education or employment, 10.2% (432) were either studying or working in education; and 7.1% (304) were healthcare workers.[@PHE2017]

@@

Figure \@ref(fig:plot-prop-pul) shows that since the 1980s the proportion of extra-pulmonary TB has increased from `r prop_extra_pil_1982` in 1982 to `r prop_extra_pil_2016` in 2016. This may be attributed to the age distribution of TB cases changing as different age groups are more likely to progress to pulmonary vs extra-pulmonary TB. It may also be related to the increase of non-UK born cases as a higher proportion of non-UK born cases have extra-pulmonary disease only (51.4%, 2,103/4,089, in 2016), compared to UK born cases (31.9%, 467/1,465, in 2016).[@PHE2017] For more details on TB in England, see Chapter \@ref(data) and the Public Health England 2017 and 2018 TB reports from which the summary data discussed above was extracted.[@PHE2017; @PHE2018]

```{r plot-prop-pul, echo = FALSE, message = FALSE, warning = FALSE, dpi = 300, fig.height = 12, fig.width = 12, fig.scap = "From 1913 until 1981 the figure shows the proportion respiratory vs. non-respiratory cases and from 1982 it shows the proportion of pulmonary vs. non-pulmonary TB.", fig.cap = "From 1913 until 1981 the figure shows the proportion respiratory vs. non-respiratory cases and from 1982 it shows the proportion of pulmonary vs. non-pulmonary TB. Figure produced using `tbinenglanddataclean`"}
plot_historic_prop_pul_tb() + theme(legend.position = "top")
```

@@

Until recently little attention has been given to any additional effects of BCG vaccination.[@Fine2005a; @Teo2006] However, there is now some evidence that BCG vaccination induces innate immune responses that may provide non-specific protection[@Kleinnijenhuis2012] and reduce all-cause neonatal mortality.[@Garly2003; @Higgins] There is some evidence that this reduction in all-cause mortality extends later in life.[@Rieckmann2016] Additionally, BCG vaccination may improve outcomes for individuals with active TB disease. TB patients with BCG scars have been found to respond better to treatment with earlier sputum smear conversion (detection of TB bacterium in a sputum sample) .[@Jeremiah2010] There is also evidence to support an association between BCG vaccination and reduced TB[@Abubakar2013] mortality. The evidence for additional effects of BCG vaccination on outcomes in individuals with notified TB, in England, is explored further in Chapter \@ref(beneficial-bcg-out).

@@

The BCG vaccine is one of the mostly widely-used vaccines worldwide, with approximately 100m doses given annually.[@The2004] However, due to the variable estimates of BCG efficacy, vaccination has been controversial since its development. The WHO recommends vaccination for all neonates as early as possible after birth in high burden settings. Vaccination in low burden settings is dependent on the country specific epidemiology of TB.[@WHO2017; @TheWorldHealthOrganization:2018va] This recommendation is based on the strong evidence that the BCG is highly protective in children,[@Rodrigues1993; @Colditz1994] whilst its effectiveness has been shown to vary with latitude when given later in life.[@Mangtani2014] Historically, different strategies have been utilized worldwide. This includes universal vaccination of those at most risk of onwards transmission and high-risk group vaccination targeting either neonates or children.[@Abubakar2013]

@@

Multiple replacement vaccines are currently in clinical trials.[@Zwerling2011a; @Schrager:2018ip] Vaccine candidates include both live and sub-unit vaccines. Many of these candidate vaccines serve as a boost to the BCG vaccine, with the BCG vaccine being administered prior to the candidate vaccine.[@Zwerling2011a] Several BCG replacements are also being trialed, both based on alternative methods of attenuating TB mycobacteria and using other approaches.[@Schrager:2018ip] However, in the short-to-midterm it is unlikely that a new vaccine will replace the BCG vaccine. This means that its optimal usage is as important as ever.
